---
title: "Effect of the algorithm on tree length"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

Effect of character correlation
===============

To measure the effect of the our inapplicable algorithm, compared to treating inapplicable tokens as missing data or an extra state, we ran a normal parsimony search on 23 <!-- @@@ --> independent matrices from a range of taxonomic groups with variable number of characters, taxa and inapplicable data.

|              |group         | taxa| characters| % NA characters| % NA cells|
|:-------------|:-------------|----:|----------:|---------------:|----------:|
|Aguado2009    |Annelids      |   76|        107|            40.2|        5.0|
|Asher2005     |Mammals       |   23|        126|            31.7|        3.9|
|Baker2009     |Angiosperms   |  205|        105|            35.2|        8.3|
|Bouchenak2010 |Angiosperms   |  818|        417|            81.5|       21.1|
|Dikow2009     |Insects       |   88|        220|             8.2|        1.6|
|Eklund2004    |Angiosperms   |   54|        131|            31.3|       10.0|
|Geisler2001   |Mammals       |   68|        186|            46.8|        2.0|
|Giles2015     |Fishes        |   78|        236|            74.2|       17.3|
|Griswold1999  |Spiders       |   43|        137|            43.8|        9.6|
|Liljeblad2008 |Insects       |   68|        308|            31.5|        6.4|
|Loconte1991   |Angiosperms   |   56|        104|            63.5|        4.1|
|Longrich2010  |Dinosaurs     |   19|        119|            26.1|        5.7|
|OMeara2013    |Mammals       |   63|        317|            44.5|        5.4|
|Rougier2012   |Mammals       |   58|        317|             2.8|        0.4|
|Sharkey2011   |Insects       |  105|        392|           100.0|       20.1|
|Sundue2010    |Pteridophytes |  177|        237|            81.9|       26.5|
|Vinther2008   |Annelids      |   23|         57|            43.9|       20.9|
|Wiens2010     |Squamates     |   64|        363|            41.9|        7.7|
|Wilson2003    |Crustaceans   |   61|        165|            43.0|       11.1|
|Wortley2006   |Angiosperms   |   37|        105|            20.0|        3.7|
|Zanol2012     |Annelids      |   80|        230|            69.1|       17.3|
|Zhu2013       |Fishes        |   75|        253|            70.8|       19.3|

From these matrices, we ran a heuristic search in PAUP* treating all inapplicable data as missing, and saving the island of _n_ trees (default PAUP* options).
The length of of the trees on this island is called _S~MP~_ the overall most parsimonious score.
All trees on this island are or the same length _S~MP~_.

The length of the trees from these island are then re-calculated by treating inapplicable as an extra state (or "new state") or by using our inapplicable algorithm.
This results in two new lists of _n_ scores (for the _n_ trees of each island) with _S~NS~_ and _S~NA~_ being the variable tree lengths for each topology under respectively the "new state" and inapplicable counting algorithms.

We then calculated how many trees were not the shortest in _S~NS~_ and _S~NA~_ (i.e. how many scores in these distributions were greater than the minimum) to measure how many trees were not discarded when treating inapplicable as missing data.
Finally, we calculated how longer (or shorter!) were the  _S~NS~_ and _S~NA~_ scores compared to the _S~MP~_ to see how both "new state" and the inapplicable algorithm modify tree length in comparison to treating inapplicable data as missing data. 

Before starting
===============

Loading the functions
-----------------------------------

Getting the functions

```{r, eval = TRUE}
source("Functions/read.tree.score.R")
source("Functions/standardise.score.R")
source("Functions/plot.scores.R")
source("Functions/test.R")
```

Loading the data
-----------------------------------

The most parsimonious islands treating inapplicable data as missing were obtained using PAUP* (v 4.0a151) with random sequences addition replicated 100 times with a maximum of 5 x 10^6 rearrangements per replicates (`hsearch addseq=random nreps=100 rearrlimit=5000000 limitperrep=yes` in PAUP).
The length of the resulting trees within the islands were then measured using PAWM with inapplicable data treated as a new state or using our inapplicable algorithm.

We can load these results using the following:

```{r, message = FALSE}
## Setting the path
path <- "Data/Scores/"

## Getting the chain names
chains <- list.files(path, pattern = ".morphy.inapplicable.count")
chains <- unlist(strsplit(chains, split = ".morphy.inapplicable.count"))

## Remove Sutton2012 (no NA)
chains <- chains[-which(chains == "Sutton2012")]

## Loading the tree scores
scores <- sapply(chains, read.tree.score, path, simplify = FALSE)

## Loading the matrices
matrices <- lapply(chains, read.matrices, path = "Data/Matrices")
```


```{r, eval = FALSE, echo = FALSE}
## Get the characteristics of each matrix

## Get my Claddis version
install_github("TGuillerme/Claddis")
library(Claddis)

## Empty matrix
matrix_characteristics <- matrix(NA, ncol = 5, nrow = length(chains))

for(matrix in 1:length(chains)) {
    ## Get the matrix
    mat_tmp <- ReadMorphNexus(paste0("Data/Matrices/", chains[[matrix]], ".nex"), gap.as.missing = FALSE)$matrix
    ## Number of taxa
    matrix_characteristics[matrix, 2] <- nrow(mat_tmp)
    ## Number of characters
    matrix_characteristics[matrix, 3] <- ncol(mat_tmp)
    ## Proportion of NA char
    matrix_characteristics[matrix, 4] <- length(which(apply(mat_tmp, 2, function(X) any(na.omit(X) == "-"))))/ncol(mat_tmp)*100
    ## Proportion of NA cells
    matrix_characteristics[matrix, 5] <- length(which(mat_tmp == "-"))/(ncol(mat_tmp)*nrow(mat_tmp))*100
}

## Convert into a data.frame
matrix_characteristics <- data.frame(matrix_characteristics)
matrix_characteristics[,2:5] <- as.numeric(matrix_characteristics[,2:5])

## Groups
matrix_characteristics[, 1] <- c("Annelids", "Mammals", "Angiosperms", "Angiosperms", "Insects", "Angiosperms", "Mammals", "Fishes", "Spiders", "Insects", "Angiosperms", "Dinosaurs", "Mammals", "Mammals", "Insects", "Pteridophytes", "Annelids", "Squamates", "Crustaceans", "Angiosperms", "Annelids", "Fishes")

## Add the rownames and colnames
colnames(matrix_characteristics) <- c("group", "taxa", "characters", "% NA characters", "% NA cells")
rownames(matrix_characteristics) <- chains


## markdown:
knitr::kable(matrix_characteristics, digits = 1)
## LaTeX
xtable::xtable(matrix_characteristics, digits = 1)
```

The scores are then standardise to reflect the proportional length of the trees compared to the shortest tree for both _S~NS~_ and _S~NA~_.

```{r}
## Normalising the score (dividing by the shortest score from each individual search)
scores_short <- mapply(standardise.score, scores, std = "short", SIMPLIFY = FALSE)
```

Getting the proportion of discarded trees under both algorithms treating inapplicable data as non missing
===============

We can then measure from these standardised scores, how many trees would have been discarded from a default PAUP search (i.e. how values in _S~NS~_ and _S~NA~_ are higher than their minimum).
Here, the proportion of discarded trees are the number of trees in the most parsimonious island (under each algorithm) that are one step or more longer than the shortest tree.

```{r}
## Getting the proportion of discarded trees
proportions <- lapply(scores_short, get.proportion)
## Only selecting these proportions for inapplicable or extra state
inapp_proportions <- unlist(lapply(proportions, `[[`, 1))
extra_proportions <- unlist(lapply(proportions, `[[`, 3))
## Combining the data into a data frame
proportions_combined <- data.frame(inapp_proportions, extra_proportions)
colnames(proportions_combined) <- c("Inapplicable", "New state")
```

We can then visualise this as follows:

```{r, fig.width=8, fig.height=8}
op <- par(bty = "n")
boxplot(proportions_combined, ylab = "Proportion of discarded (longer) trees")
par(op)
```

```{r, echo = FALSE}
## Summary table
knitr::kable(t(apply(proportions_combined, 2, summary)), digits = 2)
```

We see here a clear effect of the counting algorithm used with nearly 100% of trees discarded (the results are rounded) when not treating inapplicable data as missing data.
All these trees would have still be saved in a default analysis.
Note that, although their quantiles overlap, our inapplicable algorithm produces, on average, slightly less trees that are longer than treating inapplicable data as a new state.

How long are the trees?
-----------------------------------

This results however, only indicate how many longer trees are generated under these two counting methods, not how long each of them are.
We can compare each tree length _S~NS~_ and _S~NA~_ to their length _S~MP~_ when treating inapplicable data as missing.
This information can be added to the previous graph where in a two dimensional way: the vertical information is the same as above (proportion of discarded trees) and the horizontal gives information on, when trees are discarded, what are their minimum and maximum length compared to _S~MP~_.

```{r}
## Normalising the score (based on the MP island)
scores_MP <- mapply(standardise.score, scores, std = "MP", SIMPLIFY = FALSE)
## Getting the proportion of discarded trees
proportions_MP <- lapply(scores_MP, get.proportion)
## Only selecting these proportions for inapplicable or extra state
inapp_proportions_MP <- unlist(lapply(proportions_MP, `[[`, 1))
extra_proportions_MP <- unlist(lapply(proportions_MP, `[[`, 3))
## Combining the data into a data frame
proportions_MP <- data.frame(inapp_proportions_MP, extra_proportions_MP)
colnames(proportions_MP) <- c("Inapplicable", "New state")
## Getting the relative minimum and maximum tree length from both distributions
max_scores <- lapply(scores_MP, lapply, max)
min_scores <- lapply(scores_MP, lapply, min)
## Only selecting these proportions for inapplicable or extra state
inapp_max <- unlist(lapply(max_scores, `[[`, 1))
extra_max <- unlist(lapply(max_scores, `[[`, 3))
inapp_min <- unlist(lapply(min_scores, `[[`, 1))
extra_min <- unlist(lapply(min_scores, `[[`, 3))
## Updating the data frame (using the inapp and extra proportion from before)
proportions_combined <- data.frame(inapp_proportions, inapp_min, inapp_max,
    extra_proportions, extra_min, extra_max)
```

```{r, fig.width=8, fig.height=8}
op <- par(bty = "n")
sauronplot(proportions_combined, names = c("Inapplicable", "New state"),
    xlab = "Max/min proportional extra length to MP trees",
    ylab = "Proportion of discarded trees")
par(op)
```

Again, on the y axis is the same as above but represented as a line: the black dot is the median and the solid and dashed lines are the 50% and 95% confidence intervals respectively.
The relative additional tree length to the overall most parsimonious tree is represented on the x axis.
Each deviation represents the proportional extra length of the shortest and longest tree respectively on the left and right from the 0 line.

```{r, echo = FALSE}
## Summary table
colnames(proportions_combined)[c(2,3,5,6)] <- c("Minimum extra length (NA)", "Maximum extra length (NA)", "Minimum extra length (NS)", "Maximum extra length (NS)")
knitr::kable(t(apply(proportions_combined[,c(2,3,5,6)], 2, summary)), digits = 2)
```


Although both algorithms can generate longer trees (even for the minimum extra length), our inapplicable algorithm always generates trees less than 50% longer than the overall most parsimonious trees (apart from some trees were nearly every trees would have been discarded in the search). 
<!-- This is probably Bouchenak's matrix -->